---
title: "DynafluxR - Supplementary information"
output:
  html_document:
    df_print: paged
    toc: yes
---

```{r, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>",
  fig.width=7,
  fig.height=5
)
```

This notebook performs all computational analyses detailed in:

> **Estimating flux dynamics from time-course metabolite concentrations with dynafluxr.**
>
> S. Sokol, S. Dubiley, P. Rouan, C. Charlier, G. Lippens, P. Millard
>
> bioRxiv preprint, doi: 10.1101/XXXXXXXXXXXXX

This R Markdown notebook is available at <https://github.com/MetaSys-LISBP/dynafluxr>

Authors: Sergueï Sokol ([sokol\@insa-toulouse.fr](mailto:sokol@insa-toulouse.fr){.email}), Pierre Millard ([pierre.millard\@insa-toulouse.fr](mailto:pierre.millard@insa-toulouse.fr){.email})

Copyright INRAE, 2025

# 1. Prepare environment

Load DynafluxR and other required packages.

```{r setup}

library(dynafluxr)
library(CoRC)
library(RColorBrewer)
library(stringr)
library(vioplot)

# color generator
fun_col <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
col_6 <- brewer.pal(n=12, name='Paired')[c(2,4,6,8,10,12)]
```

Set path.

```{r}

# get current directory
wd <- getwd()

# change the path if needed
#wd <- "~/GIT/dynafluxr/sup_data"
```

# 2. Example 1: Synthetic dataset

Set paths.

```{r}

example1_dir <- file.path(wd, "toy_model")
model_dir <- file.path(example1_dir, "model")
results_dir <- file.path(example1_dir, "results")
tmp_dir <- file.path(results_dir, "tmp")
```

## 2.1. Simulation of the dataset

Simulate time-course fluxes and metabolite concentrations (20 equally time points) using a kinetic model and COPASI.

```{r}

# load kinetic model in COPASI
loadModel(file.path(model_dir, "model.cps"))

# run time course simulations (20 time points)
setTimeCourseSettings(intervals=19)
simulations <- runTimeCourse()

# save time-course concentrations
concentrations <- simulations$result[,1:9]
write.table(concentrations, file.path(results_dir, "sim_data.txt"), row.names=FALSE, sep="\t", quote=FALSE)
```

## 2.2. Dynamic flux calculation

Calculate fluxes from true (simulated) data with DynafluxR (using default values for all parameters).

```{r}

startTime <- Sys.time()

# run flux calculation
res <- cli(c("-m", file.path(results_dir, "sim_data.txt"),
             "-s", file.path(example1_dir, "network.txt")))

endTime <- Sys.time()
print(endTime - startTime)
```

Plot true vs fitted metabolite dynamics (Figure 2B).

```{r Figure_2B, fig.width = 6.3, fig.height = 4.5, fig.path='toy_model/figures/', dev=c('png', 'pdf')}

# best fit
x_sim <- res$tp
y_sim <- res$isp(res$tp)

# simulated data
x <- simulations$result[,1]
y <- simulations$result[,colnames(y_sim)]

# create colors
cols <- fun_col(ncol(y))
cols_alpha <- paste(cols, "55", sep="")

# plot simulated vs fitted data
matplot(x=x, y=y, type="p", pch=21, cex=1.5, lwd=2, col=cols, bg=cols_alpha, las=1, ylab="concentration", xlab="time")
matlines(x=x_sim, y=y_sim, lwd=2, lty=1, col=cols)
legend("topright", inset=c(0.04, 0), legend=colnames(y), col=cols, pch=21, pt.bg=cols_alpha, pt.cex=1.5, pt.lwd=2, bty="n", ncol=4)

# add legend
legend("top", inset=c(-0.08, 0), legend="best fit", col="black", lwd=2, bty="n")

# add confidence intervals
y_sim_sd_p <- res$rsp(res$tpp)
y_sim_sd_m <- res$rsp(res$tpp)
for (i in seq(ncol(y_sim))){
  polygon(x=c(res$tpp, rev(res$tpp)), y=c((y_sim_sd_p[,i]), rev(y_sim_sd_m[,i])),
          col=paste(fun_col(ncol(y))[i], "55", sep=""), border=NA)
}
```

Mean error on metabolite concentrations.

```{r}

mean_err_conc <- mean(as.matrix(abs(y-y_sim)))

mean_err_conc
```

Plot true vs fitted flux dynamics.

```{r Figure_2C, fig.width = 6.3, fig.height = 4.5, fig.path='toy_model/figures/', dev=c('png', 'pdf')}

# true flux values
x <- simulations$result[,1]
y <- simulations$result[,10:15]

# fluxes calculated with DynafluxR
x_sim <- res$tpp
y_sim <- res$vsp(res$tpp)
y_sim_sd_p <- res$vsp(res$tpp, fsd=1)
y_sim_sd_m <- res$vsp(res$tpp, fsd=-1)

# create colors
cols <- fun_col(ncol(y))
cols_alpha <- paste(cols, "55", sep="")

# plot true vs estimated fluxes
matplot(x=x, y=y, type="p", pch=21, cex=1.5, lwd=2, col=cols, bg=cols_alpha, las=1, ylab="rate", xlab="time", ylim=c(0,0.1))
matlines(x=x_sim, y=y_sim, lwd=2, lty=1, col=cols)

# add confidence intervals
for (i in seq(ncol(y_sim))){
  polygon(x=c(x_sim, rev(x_sim)), y=c((y_sim_sd_p[,i]), rev(y_sim_sd_m[,i])),
          col=paste(cols[i], "55", sep=""), border=NA)
}

# add legend
leg <- stringr::str_remove_all(stringr::str_remove_all(colnames(y), "]"), "Values\\[")
legend("topright", inset=c(-0.04, 0), legend=leg, col=cols, pch=21, pt.bg=cols_alpha, pt.cex=1.5, pt.lwd=2, bty="n", ncol=4)
legend("topright", inset=c(0.045, 0.2), legend="estimated", col="black", lwd=2, bty="n")
```

Calculate mean error on rates.

```{r}

mean_err_rates <- mean(as.matrix(abs(simulations$result[,10:15] - res$vsp(res$tp))))

mean_err_rates
```

## 2.3. Robustness analysis

Evaluate robustness of the approach with respect to number of measurements (from 5 to 20 data points) and the number of knots (from 2 to 5). Parameters are kept at default values.

```{r Figure_2D, fig.width = 6.3, fig.height = 4.5, fig.path='toy_model/figures/', dev=c('png', 'pdf')}

# number of data points (from 5 to 20)
n_points <- seq(5, 20)
# number of knots (from 2 to 5)
n_k <- seq(2, 5)
# matrix to store results
residuals_k_p <- matrix(data=NA, ncol=length(n_k), nrow=length(n_points))

for (i in n_points){
  
  # set number of data points to simulate
  setTimeCourseSettings(intervals=i-1)
  
  # run time-course simulation with COPASI
  simulations <- runTimeCourse()
  
  # save simulated metabolite dynamics (input format of DynafluxR)
  concentrations <- simulations$result[,1:9]
  fname <- file.path(tmp_dir, "tmp_1.txt")
  write.table(concentrations, file=fname, row.names=FALSE, sep="\t", quote=FALSE)
    
  # for each number of knots
  for (k in n_k){
    
    # calculate fluxes with DynafluxR
    suppressWarnings({
      res <- cli(c("-m", fname,
                   "-s", file.path(example1_dir, "network.txt"),
                   "-k", k))
    })
    
    # calculate mean flux error
    y <- simulations$result[,10:15]
    y_sim <- res$vsp(res$tp)
    residuals_k_p[i-n_points[1]+1, k-n_k[1]+1] <- mean(as.matrix(abs(y-y_sim)))
    
  }
}
```

Plot results.

```{r}

matplot(x=n_points, y=residuals_k_p, ylab="mean rate error", xlab="number of data points", las=1, pch=21, col=col_6, bg=paste0(col_6, "55"), cex=1.5, lwd=2, yaxt="n", log="y")
axis(2, at=c(1e-3, 1e-2, 1e-1), label=c(0.001, 0.01, 0.1), las=1)
axis(2, at=c(seq(0.0002, 0.0009, by=0.0001), seq(0.002, 0.009, by=0.001), seq(0.02, 0.09, by=0.01)), labels=FALSE, tcl=-0.3)

legend("topright", inset=c(0.08, 0), title="number of knots", legend=n_k, col=col_6, pch=21, pt.bg=paste(col_6, "55", sep=""), pt.cex=1.5, pt.lwd=2, bty="n", ncol=4)
```

Evaluate robustness of the approach with respect to noise (Gaussian noise, from 1 to 20 % of the highest concentration of each metabolite).

```{r Figure_2E, fig.width = 6.3, fig.height = 4.5, fig.path='toy_model/figures/', dev=c('png', 'pdf')}

# noise levels (relative to the maximal concentration of each metabolite)
l_noise <- c(0.01, 0.02, 0.05, 0.1, 0.2)

# number of datasets to generate for each noise level
n_mc <- 100

# matrix to save calculation results
residuals_mc <- matrix(data=NA, ncol=n_mc, nrow=length(l_noise))
rownames(residuals_mc) <- l_noise

# run time-course simulation with COPASI (-> true values) 
setTimeCourseSettings(intervals=99)
simulations <- runTimeCourse()
concentrations <- simulations$result[,1:9]
fluxes <- simulations$result[,10:15]

# for each noise level
for (i in seq(length(l_noise))){
  
  # run monte carlo analysis (n_mc iterations)
  for (j in seq(n_mc)){
    
    # generate noisy dataset
    sd_to_add <- apply(concentrations[-1], 2, mean)*l_noise[i]
    noise <- rnorm(prod(dim(concentrations))-nrow(concentrations), mean=0, sd=rep(sd_to_add, each=nrow(concentrations)))
    concentrations_noise <- concentrations[,-1] + matrix(noise, ncol=8)    
    concentrations_noise[concentrations_noise < 0] <- 0
    concentrations_noise <- cbind(concentrations[,1], concentrations_noise)
    colnames(concentrations_noise) <- colnames(concentrations)
    
    # save dataset in DynafluxR format
    fname <- file.path(tmp_dir, "tmp_2.txt")
    write.table(concentrations_noise, file=fname, row.names=FALSE, sep="\t", quote = FALSE)
    
    # run flux calculation with DynafluxR
    res <- cli(c("-m", fname, "-s", file.path(example1_dir, "network.txt")))
  
    # calculate mean flux error
    residuals_mc[i,j] <- mean(as.matrix(abs(fluxes - res$vsp(res$tp))))
    
  }
}
```

Plot results.

```{r}

# plot results
vioplot(x=t(residuals_mc), use.cols = TRUE, las=1, col=paste0(col_6[1], "55"), border=col_6[1], rectCol=col_6[1], lineCol=col_6[1], ylim=c(0, 0.015), cex.axis=0.85, yaxs="i")
title(xlab="measurements noise (rsd of mean metabolite concentrations)")
title(ylab="mean rate error", line=3.3)
```

# 3. Example 2: Real-time NMR analysis of a multi-enzymatic reaction system

Set paths.

```{r}

example2_dir <- file.path(wd, "multienzymatic_model")
model2_dir <- file.path(example2_dir, "kinetic_model")
```

## 3.1. Flux calculation

Calculate fluxes with DynafluxR with the following parameters:

-   skip the 24 first time points that cause instabilities (`"--skip=24"`)

-   force monotonic decrease of GLC, and monotonic increase of FBP (`"--decr", "GLC"` and `"--incr", "FBP"`)

-   apply scaling factors to FBP and F6P since only their relative concentrations are measured (`"--sf=FBP,F6P"`)

-   define standard deviation on metabolite concentrations (`"--sderr=GLC=0.05,G6P=0.05,FBP=0.1,F6P=0.05"`)

The complete list of parameters can be accessed with `cli("-h")`.

```{r}

startTime <- Sys.time()

res=cli(c("-m", file.path(example2_dir, "data.txt"),
          "-s", file.path(example2_dir, "network.txt"),
          '-a', file.path(example2_dir, "atom.tsv"),
          "-o", file.path(example2_dir, "flux_calculation"),
          "--decr", "GLC",
          "--incr", "FBP",
          "--skip=24",
          "--sf=FBP,F6P",
          "--sderr=GLC=0.05,G6P=0.05,FBP=0.1,F6P=0.05"))

endTime <- Sys.time()
```

Show calculation time.

```{r}

print(endTime - startTime)
```

Statistical evaluation of the fit.

```{r}

res$chi2tab
```

Load measured metabolite dynamics.

```{r}

data <- read.table(file.path(example2_dir, "data.txt"), header=TRUE)
data
```

Plot measured vs fitted metabolite dynamics.

```{r Figure_3B, fig.width = 6.3, fig.height = 4.5, fig.path='multienzymatic_model/figures/', dev=c('png', 'pdf')}

# plot measured metabolite dynamics
matplot(x=data[,1], y=data[,2:5], type="p", pch=21, cex=1.5, col=col_6, las=1, ylab="concentration (mM)", xlab="time (min)", lwd=1.5, bg=paste(col_6, "55", sep=""))

# add best fit
matlines(x=res$tpp, y=res$isp(res$tpp)[,c(1,3,5,6)], lwd=2.5, lty=1, col="black")

# add legend
legend("topright", inset=c(0.24, 0), legend=colnames(data[,2:5]), col=col_6, pch=21, pt.bg=paste(col_6, "55", sep=""), pt.cex=1.5, pt.lwd=2, bty="n", ncol=4)
legend("top", inset=c(0.18, 0.1), legend="best fit", col="black", lwd=2, bty="n")
```

Plot rate dynamics.

```{r Figure_3C, fig.width = 6.3, fig.height = 4.5, fig.path='multienzymatic_model/figures/', dev=c('png', 'pdf')}

# get data to plot
x_sim_full <- res$tpp
y_sim_full <- res$vsp(res$tpp)
y_sim_sd_p_full <- res$vsp(res$tpp, fsd=1)
y_sim_sd_m_full <- res$vsp(res$tpp, fsd=-1)

# plot fluxes
matplot(x=x_sim_full, y=y_sim_full, type="l", lwd=2.5, lty=1, col=col_6[1:3], las=1, ylab="rate (min-1)", xlab="time (min)")

# add confidence intervals
for (i in seq(ncol(y_sim_full))){
  polygon(x=c(x_sim_full, rev(x_sim_full)), y=c((y_sim_sd_p_full[,i]), rev(y_sim_sd_m_full[,i])), 
          col=paste(col_6[i], "55", sep=""), border=NA)
}

# add legend
leg <- stringr::str_remove_all(colnames(y_sim_full), ":")
legend("topright", inset=c(0.04, 0), legend=leg, col=col_6[1:3], lwd=2.5, pt.bg=col_6[1:3], cex=1, bty="n", ncol=4)
```

## 3.2. Detailed information on metabolite dynamics

The rate estimations allow us to isolate the time-resolved contribution of individual reactions on metabolite dynamics.

```{r Figure_3D, fig.width = 6.3, fig.height = 4.5, fig.path='multienzymatic_model/figures/', dev=c('png', 'pdf')}

x_sim <- res$tpp
y_sim <- res$vsp(res$tpp)
y_sim <- y_sim[,c("HXK:", "PGI:")]
y_sim[,"PGI:"] <- (-y_sim[,"PGI:"])

# plot fluxes
matplot(x=x_sim, y=y_sim, type="l", lwd=2.5, lty=1, col=col_6[c(1,3)], las=1, ylab="flux (mM.min-1)", xlab="time (min)")
abline(h=0)

# add legend
leg <- stringr::str_remove_all(colnames(y_sim), ":")
legend("topright", inset=c(0.04, 0), legend=leg, col=col_6[c(1,3)], lwd=2.5, pt.bg=col_6[1:3], cex=1, bty="n", ncol=4)
```

## 3.3. Evaluation of carbon balance

Plot carbon balance of consolidated dataset.

```{r Figure_3E, fig.width = 6.3, fig.height = 4.5, fig.path='multienzymatic_model/figures/', dev=c('png', 'pdf')}

# calculate carbon balance
x_sim <- res$tpp
y_sim <- res$iasp(res$tpp)
y_meas <- colSums(t(res$mf[,-1L])*res$atomlen[colnames(res$mf)[-1L]], na.rm=TRUE)

# plot measured and fitted C balance
matplot(x=res$tp, y=y_meas, type="p", pch=21, cex=1.5, lwd=1.5, bg="lightblue", col="royalblue", las=1, ylab="total carbon concentration (mM)", xlab="time (min)", ylim=c(35,40))
matlines(x=x_sim, y=y_sim, type="l", lwd=2.5, lty=1, col="black")
```

## 3.4. Prediction of FBP dynamics

```{r}

res=cli(c("-m", file.path(example2_dir, "data.txt"),
          "-s", file.path(example2_dir, "network.txt"),
          "-o", file.path(example2_dir, "prediction_FBP"),
          "--decr", "GLC",
          "--incr", "FBP",
          "--skip=24",
          "--lna=FBP",
          "--sderr=GLC=0.1,G6P=0.1,F6P=0.1"))
```

Plot rates estimated with and without FBP.

```{r Figure_3F, fig.width = 6.3, fig.height = 4.5, fig.path='multienzymatic_model/figures/', dev=c('png', 'pdf')}

# get data to plot
x_sim <- res$tpp
y_sim <- res$vsp(res$tpp)
y_sim_sd_p <- res$vsp(res$tpp, fsd=1)
y_sim_sd_m <- res$vsp(res$tpp, fsd=-1)

# plot fluxes
matplot(x=x_sim, y=y_sim, type="l", lwd=2.5, lty=1, col=col_6[1:3], las=1, ylab="rate (min-1)", xlab="time (min)")

matlines(x=x_sim_full, y=y_sim_full, type="l", lwd=2.5, lty=2, col=col_6[1:3], las=1, ylab="rate (min-1)", xlab="time (min)")

# add legend
leg <- stringr::str_remove_all(colnames(y_sim), ":")
legend("topright", inset=c(0.04, 0), legend=leg, col=col_6[1:3], pch=22, pt.bg=col_6[1:3], cex=1, bty="n", ncol=3)
legend("topright", inset=c(0.04, 0.1), lty=c(1,2), legend=c("full dataset", "without FBP"), col="black", cex=1, lwd=2, bty="n", ncol=1)
```

Predicted FBP dynamics.

```{r Figure_3G, fig.width = 6.3, fig.height = 4.5, fig.path='multienzymatic_model/figures/', dev=c('png', 'pdf')}

x <- data$Time
FBP_meas <- data$FBP
FBP_meas_scaled <- FBP_meas/max(FBP_meas)

x_pred <- res$tpp
FBP_pred <- res$isp(res$tpp)[,"FBP"]
FBP_pred_scaled <- FBP_pred/(max(FBP_pred))


plot(x=x, y=FBP_meas_scaled, type="p", cex=1.5, lwd=1.5, las=1.5, ylab="relative FBP concentration (AU)", xlab="time (min)", ylim=c(0,1.1), pch=21, col="royalblue", bg="lightblue")
lines(x=x_pred, y=FBP_pred_scaled, lwd=2.5, lty=1, col="black")
```

## 3.5. Comparison with kinetic model

Load kinetic model in COPASI.

```{r}

# load kinetic model in COPASI
loadModel(file.path(model2_dir, "kinetic_model.cps"))
```

Run parameter estimation.

```{r}

# run parameter estimation
par_model <- runParameterEstimation()

# display duration of parameter estimation
par_model$main$cpu_time_s
```

Simulate fluxes corresponding to best fit.

```{r}

# run time course simulations from best fit
simulations <- runTimeCourse()
```

Compare fluxes estimated with kinetic modeling (COPASI) and stoichiometric modeling (DynafluxR).

```{r Figure_3H, fig.width = 6.3, fig.height = 4.5, fig.path='multienzymatic_model/figures/', dev=c('png', 'pdf')}

# get fluxes from the best fit (COPASI)
x <- simulations$result[25:1025,1]
y <- simulations$result[25:1025,10:12]

# get fluxes from the best fit (DynafluxR)
x_sim <- res$tpp
y_sim <- res$vsp(res$tpp)
y_sim_sd_p <- res$vsp(res$tpp, fsd=1)
y_sim_sd_m <- res$vsp(res$tpp, fsd=-1)

# plot flux dynamics
matplot(x=x, y=y, type="l", lty=2, lwd=2.5, col=col_6[1:3], las=1, ylab="rate (min-1)", xlab="time (min)", ylim=c(0,0.05))
matlines(x=x_sim, y=y_sim, lwd=2.5, lty=1, col=col_6[1:3])

# add confidence intervals (DynafluxR)
for (i in seq(ncol(y_sim))){
  polygon(x=c(x_sim, rev(x_sim)), y=c((y_sim_sd_p[,i]), rev(y_sim_sd_m[,i])),
          col=paste(col_6[i], "55", sep=""), border=NA)
}

# add legend
leg <- stringr::str_remove_all(stringr::str_remove_all(colnames(y), "]"), "Values\\[")
legend("topright", inset=c(0.04, 0), legend=leg, col=col_6[1:3], pch=22, pt.bg=col_6[1:3], cex=1, bty="n", ncol=3)
legend("topright", inset=c(0.04, 0.1), lty=c(1,2), legend=c("stoichiometric model", "kinetic model"), col="black", cex=1, lwd=2, bty="n", ncol=1)
```
